Division and Factorization of p-Adic Polynomials

### 2016-01-09 polynomials/muldivfact.tex
Some remarks added

\todo{Write introduction.}

\noindent
{\bf Main Results.}

\noindent
{\bf Organization of the paper.}

@@

It is worth noting that tracking precision using this representation is 
rather easy. For example, if $x$ and $y$ are known with absolute (resp. 
relative) precision $N_x$ and $N_y$ respectively, one can compute the 
sum $x+y$ (resp. the product $xy$) at absolute (resp. relative) 
precision $\min(N_x,N_y)$. Computations with $p$-adic and Laurent
series are often handle this way on symbolic computation softwares.

@@

\todo{Add introduction.}
In the sequel, we use the notation $A \div B$ (resp $A \mod B$)
for the quotient (resp. the remainder) in the Euclidean division of $A$ 
by $B$.

@@

\noindent
{\bf Newton precision.}
We consider here the case where the precision is given by Newton
functions. Concretely, we pick $A, B \in K[X]$ two polynomials which 
are known at precision $O(\varphi_A)$ and $O(\varphi_B)$ respectively:
$$A = A_\app + O(\varphi_A)
\quad \text{and} \quad
B = B_\app + O(\varphi_B)$$
Here $A_\app$ and $B_\app$ are some approximations of $A$ and $B$ respectively 
and $\varphi_A$ and $\varphi_B$ denotes two Newton functions of degree 
$\deg A$ and $\deg B$ respectively.
We are interested in determining the precision on $A \mod B$ and $A
\div B$. The following proposition gives a theoretical answer under
mild assumptions.

@@

We observe several interesting properties. First of all, the gains for 
Newton precision and jagged precision always agree though one may have 
thought at first that Newton precision is weaker. Since performing
precision computations in the Newton framework is cheaper, it seems
(at least on this example) that using the jagged precision model is not
relevant.

@@

Let $t_1$ (resp. $t_2$) be the translation of vector $(0, b_\varphi
(Q_1))$ (resp. $(0, b_0(Q_2))$. It follows from the definition of
$b_\varphi$ that $\NP(Q_1)$ is a subset of $t_1(\Epi(\varphi))$ where
$\Epi(\varphi)$ denotes the epigraph of $\varphi$. Similarly $\NP(Q_2)
\subset t_2(C)$ where $C$ is the convex cone generated by the vectors 
starting from $(0,0)$ to $(0,1)$ and $(1, \lambda_0)$. Thus
$$\NP(Q_1 Q_2) \subset t_2 \circ t_1 \big(\Epi(\varphi) + C\big) =
t_2 \circ t_1\big(\Epi(\varphi)\big)$$
and \emph{c)} follows.

@@

\noindent
{\bf Lattice precision.}
The precision datum is given here by a lattice $H_P$ in $K_{\leq n{-}1}
[X]$; we shall then write:
$$P = P_\app + O(H_P)$$
where $P_\app$ is a \emph{monic} approximation of the inexact polynomial 
$P$ we want to factor. We assume from now that $H_P$ is sufficiently 
small so that the Newton polygon of $P$ is well defined. We then can
define the function:
$$F = (F_A, F_B) : P_\app + H_P \to K_{=d}[X] \times K_{=n{-}d}[X]$$
mapping a polynomial $P$ to the couple $(A^{(1)}, B^{(1)})$ obtained
from it. We set $(A_\app^{(1)}, B_\app^{(1)}) = F(P_\app)$.

@@

\begin{prop}
\label{prop:precA1}
The application $F_A : P \mapsto A^{(1)}$ is of class $C^1$ on $P_\app + H_P$
and its differential at at some point $P$ is the linear mapping:
$$dP \mapsto dA^{(1)} = (V^{(1)} \: dP) \mod A^{(1)}$$
where $(A^{(1)}, B^{(1)}) = F(P)$ and $V^{(1)}$ is the inverse of 
$B^{(1)}$ modulo $A^{(1)}$.
\end{prop}

@@

A remarkable Corollary of Proposition~\ref{prop:precA1} asserts the
optimality of Proposition~\ref{prop:Newtonprecslope} in a particular 
case.

\begin{cor}
We assume~\textbf{(H)}.
When the precision of $P$ is given by $O(\varphi_P)$ with:
$$\varphi_P = (\NF(P) + \delta)_{|[0,d-1]}
\quad \text{where $\delta \in \R$}$$
the precision of $A^{(1)}$ given by 
Proposition~\ref{prop:Newtonprecslope} is optimal.
\end{cor}

\begin{proof}
First remark that the $\delta$ defined in the statement of 
Proposition~\ref{prop:Newtonprecslope} coincide with the $\delta$
introduced in the Corollary. Define
$\varphi_{A^{(1)}} = \varphi_{|[0,d{-}1]} - \varphi(d) + \delta$.
Let $H_P$ (resp. $H_{A^{(1)}})$ be the lattice consisting of
polynomials of degree at most $n{-}1$ (resp. at most $d{-}1$)
whose Newton function is not less that $\varphi_P$ (resp. 
$\varphi_{A^{(1)}}$). We have to show that $F'_A(P_\app)(H_P) = 
H_{A^{(1)}}$. But, according to Proposition~\ref{prop:precA1}
the mapping $G : K_{\leq d{-}1}[X] \to K_{\leq n{-}1}[X]$,
$$dA^{(1)} \mapsto (B_\app^{(1)} \: dA^{(1)}) \mod A_\app^{(1)}$$
is a right inverse of $F'_A(P_\app)$. It is then enough to prove
that $G$ takes $H_{A^{(1)}}$ to $H_P$, which can be done easily
using Theorem~\ref{theo:EDivisionNP}.
\end{proof}

>>>>

@@

It worths be noting that tracking precision using this representation is 
rather easy. For example, if $x$ and $y$ are known with absolute (resp. 
relative) precision $N_x$ and $N_y$ respectively, one can compute the 
sum $x+y$ (resp. the product $xy$) at absolute (resp. relative) 
precision $\min(N_x,N_y)$. Computations with $p$-adic and Laurent
series are often handle this way on symbolic computation softwares.

@@

\todo{Add introduction.}
In the sequel, we shall will the notation $A \div B$ (resp $A \mod B$)
for the quotient (resp. the remainder) in the Euclidean division of $A$ 
by $B$.

@@

\noindent
{\bf Newton precision.}
We consider here the case where the precision is given by Newton
functions. Concretely, we pick $A, B \in K[X]$ two polynomials which 
are known at precision $O(\varphi_A)$ and $O(\varphi_B)$ respectively:
$$A = A_\app + O(\varphi_A)
\quad \text{and} \quad
B = B_\app + O(\varphi_B)$$
Here $A_\app$ and $B_\app$ are some approximations of $A$ and $B$ respectively 
and $\varphi_A$ and $\varphi_B$ denotes two Newton functions of degree 
$\deg A$ and $\deg B$ respectively.
We are interesting in determining the precision on $A \mod B$ and $A
\div B$. The following proposition gives a theoretical answer under
mild assumptions.

@@

We observe several interesting properties. First of all, the gains for 
Newton precision and jagged precision always agree though one may has 
thought at first that Newton precision is weaker. Since performing
precision computations in the Newton framework is cheaper, it seems
(at least on this example) that using the jagged precision model is not
relevant.

@@

Let $t_1$ (resp. $t_2$) be the translation of vector $(0, b_\varphi
(Q_1))$ (resp. $(0, b_0(Q_2))$. It follows from the definition of
$b_\varphi$ that $\NP(Q_1)$ is a subset of $t_1(\Epi(\varphi))$ where
$\Epi(\varphi)$ denotes the epigraph of $\varphi$. Similarly $\NP(Q_2)
\subset t_2(C)$ where $C$ is the convex cone generated by the vectors
$(0,1)$ and $(1, \lambda_0)$. Thus
$$\NP(Q) \subset t_2 \circ t_1 \big(\Epi(\varphi) + C\big) =
t_2 \circ t_1\big(\Epi(\varphi)\big)$$
and \emph{c)} follows.

@@

\noindent
{\bf Lattice precision.}
The precision datum is given here by a lattice $H_P$ in $K_{\leq n{-}1}
[X]$; we shall then write:
$$P = P_\app + O(H_P)$$
where $P_\app$ is a \emph{monic} approximation of the inexact polynomial 
$P$ we want to factor. We assume from now that $H_P$ is sufficiently 
small so that the Newton polygon of $P$ is well defined. We then can
define the function:
$$F : P_\app + H_P \to K_{=d}[X] \times K_{=n{-}d}[X]$$
mapping a polynomial $P$ to the couple $(A^{(1)}, B^{(1)})$ obtained
from it. We set $(A_\app^{(1)}, B_\app^{(1)}) = F(P_\app)$.

@@

\begin{prop}
\label{prop:precA1}
The application $P \mapsto A^{(1)}$ is of class $C^1$ on $P_\app + H_P$
and its differential at at some point $P$ is the linear mapping:
$$dP \mapsto dA^{(1)} = (V^{(1)} \: dP) \mod A^{(1)}$$
where $(A^{(1)}, B^{(1)}) = F(P)$ and $V^{(1)}$ is the inverse of 
$B^{(1)}$ modulo $A^{(1)}$.
\end{prop}

@@
